\documentclass{llncs}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{url,color}

\begin{document}
\title{The Index $j$ in RC4 is not Pseudo-random}
\author{}
\institute{}
\maketitle

\begin{abstract}
In this paper we provide several theoretical evidences that the pseudo-random index $j$ of RC4 is indeed not pseudo-random. 
First we show that in long term $\Pr(j = i+1) = \frac{1}{N} - \frac{1}{N^2}$, instead 
of the random association $\frac{1}{N}$ and this happens for the non-existence of the condition $S[i] = 1 \mbox{ and } j = i+1$
that is mandatory for the non-existence of the Finney cycle. Further we also identify several results on non-existence of certain sequences 
of $j$. 
\end{abstract}
{\bf Keywords:} RC4, Non-randomness, Pseudo-random Index, Stream Cipher, Cryptography.

\section{Introduction}
As we all know, there are many results related to non-randomness of RC4 that received the attention in flagship level
cryptology conferences and journals (see for example~\cite{asia,joc,euro} and the references therein). Even after intense research 
for more than three decades on a few lines of RC4 algorithm, we are still amazed with new discoveries in this area of research. 
As we are presenting a short note, we assume that the reader is aware of RC4 algorithm. Still let us present the algorithm briefly.

In RC4, there is a $N = 256$ length array of 8-bit integers $0$ to $N-1$, that works as a permutation. There is also an $l$ length array
of bytes $K$, where $l$ may vary from 5 to 32, depending on the key length. There are also two bytes $i, j$, where $i$ is the deterministic
index that increases by 1 in each step and $j$ is updated in a manner so that it behaves pseudo-randomly.
The Key Scheduling Algorithm (KSA) of RC4 is as follows:
\begin{itemize}
\item $j = 0$; for $i = 0$ to $N-1$: $S[i] = i$; 
\item for $i = 0$ to $N-1$:
\begin{itemize}
\item[] $j = j + S[i] + K[i \bmod l]$; swap($S[i], S[j]$);
\end{itemize}
\end{itemize}
Next the pseudo-random bytes $z$ are generated during the Pseudo Random Generator Algorithm (PRGA) as follows:
\begin{itemize}
\item $i = j = 0$;
\item for $i = 0$ to $N-1$:
\begin{itemize}
\item[] $i = i+1$; $j = j + S[i]$; swap($S[i], S[j]$); $z = S[S[i]+S[j]]$; 
\end{itemize}
\end{itemize}
Note that all the additions here are modulo $N$.

\section{Non-Randomness due to non-existence of Finney cycle}
While there is long term suspicion that there could be problems with the psudo-randomness of $j$, till very recently it could not be 
observed or reported. In fact, in~\cite[Section 3.4]{joc}, non-randomness of $j$ has been studied for initial rounds and it has been
commented that the distribution of $j$ is almost uniform for higher rounds. Thus, to date, no long term pseudo-randomness 
of the index $j$ has been reported.

It has been observed by Finney~\cite{finney} that if $S[i] = 1$ and $j = i+1$, then RC4 lands into a short cycle of length $N(N-1)$.
Fortunately (or knowing this very well), the design of RC4 by Rivest
considers the initialization of RC4 PRGA as $i = j = 0$. Thus, during RC4 PGRA, the Finney cycle cannot occur, i.e., if
$\Pr(S[i] = 1)$, then $\Pr(j = i+1) = 0$. This provides
the non-randomness in $j$.

\begin{theorem}
\label{th1}
During RC4 PRGA, $\Pr(j = i+1) = \frac{1}{N} - \frac{1}{N^2}$, under certain usual assumptions.
\end{theorem} 
\begin{proof}
We have
\begin{eqnarray*}
\Pr(j = i+1) & = & \Pr(j = i+1, S[i] = 1) + \Pr(j = i+1, S[i] \neq 1)\\
             & = & 0 + \Pr(j = i+1 | S[i] \neq 1) \cdot \Pr(S[i] \neq 1)\\
             & = & \frac{1}{N} \cdot (1 - \frac{1}{N}) = \frac{1}{N} - \frac{1}{N^2}.
\end{eqnarray*} 
Here we consider $\Pr(j = i+1 | S[i] \neq 1) = \frac{1}{N}$ under usual 
randomness assumption (it has been checked by experiments too). 
Further, considering $S$ as a random permutation, we get $\Pr(S[i] \neq 1) = 1 - \frac{1}{N}$. \qed
\end{proof}

In fact, one can sharpen this result slightly by using Glimpse theorem as follows. Though it happens generally once out of $N$ rounds
during the PRGA.
\begin{corollary}
During RC4 PRGA, $\Pr(j = i+1 | i = z+1) = \frac{1}{N} - \frac{2}{N^2} + \frac{1}{N^3}$.
\end{corollary}
\begin{proof}
We refer to Glimpse theorem~\cite{jenkins} that says, $\Pr(S[j] = i - z) = \frac{2}{N} - \frac{1}{N^2}$ after the swap of
$S[i]$ and $S[j]$. Consider the situation when $S[i] = 1$ before the swap. That means $S[j] = 1$ after the swap. 
Thus, $\Pr(S[i] = 1 | i = z+1) = \frac{2}{N} - \frac{1}{N^2}$. Hence, we have the following:
\begin{eqnarray*}
\Pr(j = i+1 | i = z+1) & = & \Pr(j = i+1, S[i] = 1 | i = z+1)\\
                       &   & + \Pr(j = i+1, S[i] \neq 1 | i = z+1)\\
             & = & 0\\ 
             &   & + \Pr(j = i+1 | S[i] \neq 1, i = z+1)\\
             &   & \cdot \Pr(S[i] \neq 1 | i = z+1)\\
             & = & \frac{1}{N} \cdot (1 - \frac{2}{N} + \frac{1}{N^2}) = \frac{1}{N} - \frac{2}{N^2} + \frac{1}{N^3}.
\end{eqnarray*} 
We consider the usual assumptions as in Theorem~\ref{th1}. \qed
\end{proof}

Since we make a few assumptions, it is important to validate the results and
the experimental data indeed supports the theoretical claims mentioned above. 

\section{Non-existant sequences of $j$ over several rounds}

\section{Conclusion}
Rewrite

The pseudo-randomness of the index $j$ in RC4 has been an open question for quite some time. In this note we show that
$j$ is indeed not pseudo-random in long term evolution of RC4 PRGA where we consider $S$ as a pseudo-random permutation. 
To the best of our knowledge, this result has not been noted earlier. The implication of this result could be interesting 
to obtain further non-randomness in the evolution of RC4. Moreover, the result may be utilized to obtain additional biases 
at the initial stage of RC4 PRGA where the permutation $S$ has certain non-randomness.  

\begin{thebibliography}{10}

\bibitem{finney}
H. Finney. 
An RC4 cycle that can't happen. Post in sci.crypt, September 1994.

\bibitem{jenkins}
R. J. Jenkins. ISAAC and RC4. 1996. 
Available at \url{http://burtleburtle.net/bob/rand/isaac.html} 
[last accessed on October 25, 2015].

\bibitem{asia} 
K. G. Paterson, B. Poettering and J. C. N. Schuldt. 
Big Bias Hunting in Amazonia: Large-scale Computation and Exploitation of RC4 Biases. ASIACRYPT 2014. 
LNCS, Part 1, pp. 398--419, Vol. 8873, 2014.

\bibitem{joc} 
S. SenGupta, S. Maitra, G. Paul, S. Sarkar.
(Non--)Random Sequences from (Non--)Random Permutations -- Analysis of RC4 stream cipher.
Journal of Cryptology, 27(1):67--108, 2014

\bibitem{euro}
P. Sepehrdad, S. Vaudenay, and M. Vuagnoux. 
Statistical Attack on RC4 - Distinguishing WPA.
EUROCRYPT 2011. LNCS pp. 343--363, Vol. 6632, 2011.

\end{thebibliography}
\end{document}
